---
title: "Code for Section 5: Mapping Zinc in Teff (Ethiopia)"
authors: "Hangcen Zou and Debashis Mondal"
date: "`r Sys.Date()`"
output: html_document
fontsize: 11pt
geometry: margin=1in
---

# Overview

This document provides reproducible code for the analysis presented in Section 5 of the main manuscript. The workflow includes:

1. Data loading and preprocessing
2. Construction of a grid graph for kriging locations
3. Alignment of observed samples with the grid
4. Construction of neighborhood structure
5. Construction of adjacency and Laplacian matrices
6. Construction of the F  matrix, including treatment of close pairs
7. REML parameter estimation via h-likelihood
8. Results and output files

---

# Libraries
```{r libraries, message=FALSE, warning=FALSE}
library(raster)
library(igraph)
library(dplyr)
library(Matrix)
library(sparseinv)
library(readxl)
library(fields)
library(MASS)
library(spam)
library(LatticeKrig)
```


# 1. Load Data
```{r}
# environmental variables
env_e <- read_excel("../Data/Ethiopia/Ethiopia_data.xlsx")
# soil variables
soil_e <- read_excel("../Data/Ethiopia/GeoN_Ethiopia_pH_OCdata.xlsx")

# combine info
match_idx <- match(env_e$SoilSampleID, soil_e$SoilSampleID)
covariate_e <- env_e
covariate_e$ph  <- soil_e$pH_Wa[match_idx]
covariate_e$org <- soil_e$Org_pct[match_idx]

# remove samples with missing ph/org
covariate_e <- covariate_e[!is.na(covariate_e$ph),]
covariate_e <- covariate_e[!is.na(covariate_e$org),]
# remove samples with missing zinc value
covariate_e <- covariate_e[covariate_e$Zn <= 77777,]

# number of actual samples
n0 <- nrow(covariate_e)

# Ethiopia kriging point locations
target.e <- read.csv("../Data/Ethiopia/Ethiopia_map.csv", header = TRUE)

# grid length 
grid_width <- 0.013
```


# 2. Grid Graph Construction
```{r}
# use raster to get necessary info
ras_e <- rasterFromXYZ(target.e)
# number of latitude grids
d_lat <- dim(ras_e)[1]
# number of longitude grids
d_long <- dim(ras_e)[2]
# total number of grids
d_all <- d_lat*d_long

# order the kriging locations
target_e <- target.e[order(target.e$Lat,target.e$Long),]
# longitude grid index
target_long_idx <- as.integer((target_e$Long-extent(ras_e)[1])/grid_width)+1
# latitude grid index
target_lat_idx <- as.integer((target_e$Lat-extent(ras_e)[3])/grid_width)+1
# combined grid index for each location
Eth_index <- (target_long_idx-1)*d_lat + target_lat_idx
```


# 3. Align Observations to Grid
```{r}
# order the actual locations
data_e <- covariate_e[order(covariate_e$Latitude, covariate_e$Longitude),]
# longitude grid index
data_long_idx <- as.integer((data_e$Longitude-extent(ras_e)[1])/grid_width)+1
# latitude grid index
data_lat_idx <- as.integer((data_e$Latitude-extent(ras_e)[3])/grid_width)+1
# combined grid index for each location
data_index <- (data_long_idx-1)*d_lat + data_lat_idx

# check actual locations which fall outside the kriging frame
which(as.numeric(data_index %in% Eth_index)==0)


#------------ add additional grid ------------#

add <- which(as.numeric(data_index %in% Eth_index)==0)
# in order to use rbind, need to first match column names
add_data <- data_e[add,c("Longitude","Latitude")]
colnames(add_data) <- c("Long", "Lat")
# can add additional points 
target_e <- rbind(target_e,add_data)

# find the new ordering and indexing
target_e <- target_e[order(target_e$Lat,target_e$Long),]
# total number of grids
n <- nrow(target_e)
# longitude grid index
target_long_idx <- as.integer((target_e$Long-extent(ras_e)[1])/grid_width)+1
# latitude grid index
target_lat_idx <- as.integer((target_e$Lat-extent(ras_e)[3])/grid_width)+1
# combined grid index for each location
Eth_index <- (target_long_idx-1)*d_lat + target_lat_idx
```


# 4. Neighborhood Structure
```{r}
#------------ 8-neighbor graph ------------# 

# create lattice with 4 nearest neighbors
rr <- make_lattice((c(d_lat,d_long)), nei=1)
# add diagonal neighbors 
# e.g. a1 and b1 indicates [2,2]'s upper-left neighbor/edge, by symmetry [1,1]'s lower-right
# create matrix indicating grid index (column first)
a1 <- as.numeric(matrix(1:d_all,d_lat,d_long)[1:(d_lat-1),1:(d_long-1)])
b1 <- as.numeric(matrix(1:d_all,d_lat,d_long)[2:d_lat,2:d_long])
a2 <- as.numeric(matrix(1:d_all,d_lat,d_long)[1:(d_lat-1),2:d_long])
b2 <- as.numeric(matrix(1:d_all,d_lat,d_long)[2:d_lat,1:(d_long-1)])
a0 <- c(a1,a2)
b0 <- c(b1,b2)
# every two numbers indicates an edge between the corresponding nodes/grids (check doc of add_edges)
e_seq <- as.numeric(t(cbind(a0,b0))) 
# add weight to the original 4 neighbors/edges
g <- set_edge_attr(rr, "weight", value=1)
# add 4 new diagonal neighbors/edges
g <- add_edges(rr, e_seq , weight =1)



#------------ 12-neighbor graph ------------#

# create lattice with 12 nearest neighbors
rr <- make_lattice((c(d_lat,d_long)), nei=2)
# add weight to the 12 neighbors/edges
g <- set_edge_attr(rr, "weight", value=1)
```


# 5. Adjacency and Laplacian
```{r}
#------------ adjacency matrix ------------#
# adjacency matrix of the whole 
A <- as_adjacency_matrix(g, sparse=TRUE)
# adjacency matrix of kriging grids
adj_matrix <- A[Eth_index,Eth_index]

# check if the graph is connected
g_adj_matrix <- graph_from_adjacency_matrix(adj_matrix)
is_connected(g_adj_matrix)

#------------ Laplacian matrix ------------#
# Laplacian matrix
Q <- Diagonal(n, rowSums(adj_matrix))-adj_matrix

# normalize the degree of each vertex 
Q <- Q/mean(as.numeric(diag(Q)))
# add perturbation
Q <- Q + 1e-9 * Diagonal(n)
# LDU Cholesky decomposition 
K <- Cholesky(Q, perm = TRUE, super = FALSE, LDL = TRUE)
# expand the original compact representation 
K0 <- expand2(K, LDL = TRUE)
```


# 6. F Matrix and Close Pair Handling
```{r}
#------------ F matrix for all ------------#
# find index of matched grids
data_idx <- match(data_index,Eth_index)
Fmatrix <- Matrix(0, nrow=n0, ncol=n, sparse=TRUE)
for(k in 1:n0)  Fmatrix[k, data_idx[k]] <- 1


#------------ F matrix for teff ------------#
# index of teff samples in data (not grid)
teff_idx_data_all <- which(data_e$Crop=="Teff")
# teff data
teff_data_e_all <- data_e[teff_idx_data_all,]

# temporary Fmatrix corresponding to Teff
F_teff <- Fmatrix[teff_idx_data_all,]


#------------ deal with close pairs ------------#
# check if have close neighbors (falling into the same grid)
close_neighbor_grid_idx <- which(colSums(F_teff)>1)

# one of the two close neighbors should be removed
remove_idx <- c()

for(k in 1:length(close_neighbor_grid_idx)){
  close_neighbor_idx <- which(F_teff[,close_neighbor_grid_idx[k]]==1)
  # compute data mean and replace the first one
  teff_data_e_all[close_neighbor_idx[1],5:17] <- as.list(colMeans(teff_data_e_all[close_neighbor_idx,5:17]))
  # second one would be removed
  remove_idx[k] <- close_neighbor_idx[2]
}

# remove neighbors from F matrix and data
F_teff <- F_teff[-remove_idx,]
teff_data_e <- teff_data_e_all[-remove_idx,]
teff_idx_data <- teff_idx_data_all[-remove_idx]

# final number of teff samples
n0_teff <- nrow(F_teff)
```


# 7. REML Parameter Estimation via h-likelihood
```{r}
# choose a trace metal to analyze
y <- as.matrix(teff_data_e[,"Zn"])

# design matrix x (subtract mean)
x <- apply(teff_data_e[,c("Bio1","Bio12","ph")], 2, function(x) x-mean(x))

# covariates of gamma glm
z1 <- c(rep(1,n0_teff), rep(0,n))
z2 <- c(rep(0,n0_teff), diag(K0$D))

# measure convergence
ep <- 1
# number of iterations
dummy <- 1
# initial parameter value
lam0 <- 5
lam1 <- 2
est_lam <- c(lam0, lam1)

# keep track of valid locations
valid_err_idx <- seq(1,n0_teff+n,1)
valid_H_idx <- seq(1,n0_teff+n,1)

# keep track of time used
start_time <- Sys.time()

# h-likelihood iteration
while(ep>.00001){
  # print iteration
  print(dummy)
  
  # update parameters
  lam0 <- est_lam[1]
  lam1 <- est_lam[2]
  
  #------------ augmented model ------------#
  
  # precision matrix with parameters
  precision_matrix <- lam0*t(F_teff)%*%F_teff + lam1*Q
  
  # component of BLUP
  ZgammaZ <- rbind(cbind(lam0*t(x)%*%x,lam0*t(x)%*%F_teff), cbind(lam0*t(F_teff)%*%x,precision_matrix))
  
  # BLUP
  be <- lam0*solve(ZgammaZ,rbind(t(x),t(F_teff))%*%y,sparse=T)
  beta_hat <- be[1:3]
  psi_hat <- be[-(1:3)]
  
  # error
  err <- rbind(y-x%*%beta_hat-F_teff%*%psi_hat, -K0$L1.%*%K0$P1%*%psi_hat)
  
  
  #------------ gamma glm ------------#
  
  # permutation matrix
  permutation <- K0$P1.*1
  # selected inverse of the precision matrix (match permutation with Cholesky decomposition)
  TD_inv_precision <- Takahashi_Davis(Q=precision_matrix, P=permutation, return_perm_chol=1)
  
  # component of hat matrix
  inv_pre_Fx <- solve(precision_matrix,lam0*t(F_teff)%*%x,sparse=T)
  A <- lam0*t(x)%*%x - lam0*t(x)%*%F_teff%*%inv_pre_Fx
  
  # H11
  H11 <- diag(lam0*(x-F_teff%*%inv_pre_Fx) %*% solve(A) %*% (t(x)-t(inv_pre_Fx)%*%t(F_teff)) + lam0*F_teff%*%TD_inv_precision$S%*%t(F_teff))
  # H22
  C <- K0$L1.%*%K0$P1%*%inv_pre_Fx%*%t(chol(solve(A)))
  H22 <- lam1*rowSums(as.matrix(C^2))%*%K0$D + lam1*rowSums((K0$L1.%*%K0$P1%*%TD_inv_precision$S) * (K0$L1.%*%K0$P1) * diag(K0$D))
  
  # diagonal of hat matrix
  H <- c(H11,as.array(H22))
  
  # response of gamma glm
  res <- as.matrix(err^2/(1-H))
  # valid locations for gamma glm
  valid_err_idx <- intersect(valid_err_idx, which(err>10^(-6))) 
  valid_H_idx <- intersect(valid_H_idx, which(H<1-10^(-6))) 
  valid_idx <- intersect(valid_err_idx, valid_H_idx)
  # gamma glm with default inverse link
  lam.glm <- glm( res[valid_idx]~-1+z1[valid_idx]+z2[valid_idx],family=Gamma(), weights=(1-H[valid_idx]))
  # estimation of lam0 and lam1
  lam.new <- as.numeric(lam.glm$coeff)
  
  # updating
  ep <- sum(abs(lam.new-est_lam)/est_lam)
  print(lam.new)
  est_lam <- lam.new
  dummy <- dummy+1
}

Sys.time()-start_time
```


# 8. Results
```{r}
#------------ Results ------------# 

# BLUP
beta_hat
psi_hat
# estimated precision parameters
est_lam

# standard error of beta 
se_beta <- sqrt(diag(Takahashi_Davis(ZgammaZ))[1:3])
# standard error of lambda 
summary(lam.glm)

# standard error of prediction 
se_predict <- sqrt(diag(TD_inv_precision$S))

# log REML likelihood
2-lam.glm$aic/2


#------------Output Files------------# 

dir_path <- file.path("..", "Output")

write(psi_hat, file = file.path(dir_path, "ethiopia_psi_hat.txt"), ncolumns = 1)
write(se_beta, file = file.path(dir_path, "ethiopia_se_beta.txt"), ncolumns = 1)

location <- cbind(target_long_idx,target_lat_idx)

write.table(location, file = file.path(dir_path, "location_e.txt"),row.names = FALSE,col.names = TRUE)
write.table(teff_data_e_all, file = file.path(dir_path, "teff_e.txt"),row.names = FALSE,col.names = TRUE)
```

