---
title: "Code for All Figures in Section 5"
authors: "Hangcen Zou and Debashis Mondal"
date: "`r Sys.Date()`"
output: html_document
fontsize: 11pt
geometry: margin=1in
---

# Overview

This document provides reproducible code for the figures presented in the main manuscript, including:

1. Map of Africa
2. Map of the sampling domain in Ethiopia
3. Map of the sampling domain in Malawi
4. BLUP of the spatial effect
5. Level plot of the BLUP
6. Log standard error of predictions
7. 95% confidence interval plot of covariate effects


# Libraries
```{r}
library(raster)
library(readxl)
library(fields)
library(rnaturalearth)
library(ggplot2)
```


# Load Data and Output
```{r}
#------------ Ethiopia ------------# 

target.e <- read.csv("../Data/Ethiopia/Ethiopia_map.csv", header = TRUE) # Data
teff_e <- read.csv("../Output/teff_e.txt", sep="") # Output
location_e <- read.csv("../Output/location_e.txt", sep="") # Output

ras_e <- rasterFromXYZ(target.e)
grid_width <- 0.013
ras_e_longitude <- seq(extent(ras_e)[1],extent(ras_e)[2],grid_width)
ras_e_latitude <- seq(extent(ras_e)[3],extent(ras_e)[4],grid_width)
z <- matrix(NA, nrow=length(ras_e_longitude), ncol=length(ras_e_latitude))


#------------ Malawi ------------# 

maize_m <- read.csv("../Output/maize_m.txt", sep="") # Output
location_m <- read.csv("../Output/location_m.txt", sep="") # Output


#------------ h-likelihood results ------------# 
# should change to the desired result file name and corresponding path

psi_hat <- unlist(read.table("../Output/ethiopia8_psi_hat.txt", quote="\"", comment.char=""))
se_predict <- unlist(read.table("../Output/ethiopia8_se_predict.txt", quote="\"", comment.char=""))
```


# 1. Map of Africa
```{r}
africa <- ne_countries(continent = "Africa", returnclass = "sf")

# highlight Ethiopia and Malawi 
# Create a new column to specify the fill color
africa$fill_color <- ifelse(africa$name %in% c("Malawi"), "black", "white")
africa$fill_color[which(africa$name=="Ethiopia")] <- "black"
# Plot the map with specified fill and border colors
plot(africa["fill_color"], col = africa$fill_color, border = "black", main="")

# create coordinates for the map
# refer to https://en.wikipedia.org/wiki/Extreme_points_of_Africa
par(fig = c(0.15, 0.85, 0.15, 0.85))
plot(africa["fill_color"], col = africa$fill_color, border = "black", main="",cex.lab = 1.5,cex.axis = 1.3)
# Overlay a new plot with a different coordinate system
par(new = TRUE, fig = c(0.13, 0.8, 0, 0.95))
plot(x=0, y=0, xlim=c(-17.3,51.3), ylim=c(-34.5,37.2), xlab = "Longitude", ylab = "Latitude", col="white",cex.lab = 1.5,cex.axis = 1.3)
# back to default setting
par(new = FALSE, fig = c(0, 1, 0, 1))
```


# 2. Map of Sample Domain in Ethiopia
```{r}
z[cbind(location_e$Longitude,location_e$Latitude)] <- 0

# sample domain
image(ras_e_longitude,ras_e_latitude,z,col="#eeeeee",xlab="Longitude",ylab="Latitue",cex.lab = 1.5,cex.axis = 1.3)
# sampling location
points(teff_e$Longitude,teff_e$Latitude,pch=16,cex=0.4)
```


# 3. Map of Sample Domain in Malawi
```{r}
# sample domain
plot(location_m$Longitude,location_m$Latitude,pch=20,cex=0.5,col="#eeeeee",xlab='Longitude',ylab='Latitude',cex.lab = 1.5,cex.axis = 1.3)
# sampling location
points(maize_m$Longitude,maize_m$Latitude,pch=16,cex=0.4)
```


# 4. BLUP of Spatial Effect 
```{r}
#------------ Ethiopia ------------# 

# BLUP values
z[cbind(location_e$Longitude,location_e$Latitude)] <- psi_hat
# plot
image.plot(ras_e_longitude,ras_e_latitude,z,col=gray.colors(100,start=1,end=0),cex.lab=1.5,cex.axis=1.3,legend.width=0.8)


#------------ Malawi ------------# 

# create a function to generate a continuous color palette (light - small value)
rbPal <- colorRampPalette(c('white','black'))
# adds a column of color values based on the BLUP values
Col <- rbPal(100)[as.numeric(cut(psi_hat,breaks = 100))]
# plot
plot(location_m$Longitude,location_m$Latitude,pch=20,cex=0.1,col=Col,xlab='Longitude',ylab='Latitude',cex.lab = 1.5,cex.axis = 1.3)
# add legend
info<- setupLegend()
addLegend(info, col=rbPal(100), zlim = c(min(psi_hat),max(psi_hat)))

```


# 5. Level Plot of BLUP 
```{r}
#------------ Ethiopia ------------# 

# number of levels
n_colors <- 4
# plot
image.plot(ras_e_longitude,ras_e_latitude,z,col=gray.colors(n_colors,start=0.75,end=0),cex.lab = 1.5,cex.axis = 1.3,legend.width=0.8)


#------------ Malawi ------------# 

# number of levels
n_colors <- 4
# create a function to generate a continuous color palette (light - small value)
rbPal <- colorRampPalette(c('lightgray','black'))
# adds a column of color values based on the data values
Col <- rbPal(n_colors)[as.numeric(cut(psi_hat,breaks = n_colors))]
# plot
plot(location_m$Longitude,location_m$Latitude,pch=20,cex=0.1,col=Col,xlab='Longitude',ylab='Latitude',cex.lab = 1.5,cex.axis = 1.3)
# add legend
info<- setupLegend()
addLegend(info, col=rbPal(n_colors), zlim = c(min(psi_hat),max(psi_hat)))

```


# 6. Log Standard Error of Predictions
```{r}
#------------ Ethiopia ------------# 

# log se values
z[cbind(location_e$Longitude,location_e$Latitude)] <- log(se_predict)
# plot
image.plot(ras_e_longitude,ras_e_latitude,z,col=gray.colors(100,start=1,end=0),cex.lab = 1.5,cex.axis = 1.3,legend.width=0.8)


#------------ Malawi ------------# 

# create a function to generate a continuous color palette (light - small value)
rbPal <- colorRampPalette(c('white','black'))
# adds a column of color values based on the data values
Col <- rbPal(100)[as.numeric(cut(log(se_predict),breaks = 100))]
# plot
plot(location_m$Longitude,location_m$Latitude,pch=20,cex=0.1,col=Col,xlab='Longitude',ylab='Latitude',cex.lab = 1.5,cex.axis = 1.3)
# add legend
info<- setupLegend()
addLegend(info, col=rbPal(100), zlim = c(min(log(se_predict)),max(log(se_predict))))

```


# 7. 95% CI Plot of Covariate Effects
```{r}
# should change values in df based on actual results

#--------------- Ethiopia ---------------#

df <- data.frame(
  model = rep(c("8-neighbor", "12-neighbor"), each = 3),
  parameter = rep(c("beta_temp", "beta_prcp", "beta_ph"), 2),
  estimate = c(0.05972, -0.00334, -2.03392,
               0.05156, -0.00278, -1.78938),
  se = c(0.02002, 0.00118, 0.44542,
         0.02019, 0.00125, 0.43629)
)

df$model <- factor(df$model, levels = c("8-neighbor", "12-neighbor"))

# compute 95% confidence intervals
df <- df %>%
  mutate(
    lower = estimate - 1.96 * se,
    upper = estimate + 1.96 * se
  )

# create a categorical order and readable parameter labels 
df$parameter <- factor(df$parameter, 
                       levels = c("beta_temp", "beta_prcp", "beta_ph"),
                       labels = c(expression(Temperature), 
                                  expression(Precipitation), expression(pH)))

# plot
ggplot(df, aes(x = parameter, y = estimate, 
               color = model, group = model)) +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  geom_errorbar(aes(ymin = lower, ymax = upper),
                position = position_dodge(width = 0.5),
                width = 0.2, size = 0.8) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "gray40") +
  scale_color_manual(values = c("black", "gray60"),
                     name = "Neighborhood Model") +
  labs(
    x = NULL,
    y = "Estimated covariate effect"
  ) +
  theme_minimal(base_size = 20) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )



#--------------- Malawi ---------------#

df <- data.frame(
  model = rep(c("small", "large"), each = 2),
  parameter = rep(c("beta_temp", "beta_tim"), 2),
  estimate = c(0.08165, -0.12931,
               0.08230, -0.12839),
  se = c(0.00883, 0.07805,
         0.00854, 0.07737)
)

# create a categorical order and readable parameter labels 
df$parameter <- factor(df$parameter, 
                       levels = c("beta_temp", "beta_tim"),
                       labels = c(expression(Temperature), 
                                  expression(TIM)))

df$model <- factor(df$model, levels = c("small", "large"))

# compute 95% confidence intervals
df <- df %>%
  mutate(
    lower = estimate - 1.96 * se,
    upper = estimate + 1.96 * se
  )

# plot
ggplot(df, aes(x = parameter, y = estimate, 
               color = model, group = model)) +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  geom_errorbar(aes(ymin = lower, ymax = upper),
                position = position_dodge(width = 0.5),
                width = 0.2, size = 0.8) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "gray40") +
  scale_color_manual(values = c("black", "gray60"),
                     name = "Neighborhood Model") +
  labs(
    x = NULL,
    y = "Estimated covariate effect"
  ) +
  theme_minimal(base_size = 20) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )
```


