---
title: "Code for Section 5: Mapping Zinc in Maize (Malawi)"
authors: "Hangcen Zou and Debashis Mondal
date: "`r Sys.Date()`"
output: html_document
fontsize: 11pt
geometry: margin=1in
---

# Overview

This document provides reproducible code for the analysis presented in the main manuscript. The workflow includes:

1. Data loading and preprocessing
2. Construction of proximity (adjacency) matrices
3. Construction of the Laplacian matrix
4. Construction of the F matrix
5. REML parameter estimation via h-likelihood
6. Results and output files

---

# Libraries
```{r libraries, message=FALSE, warning=FALSE}
library(raster)
library(igraph)
library(dplyr)
library(Matrix)
library(sparseinv)
library(readxl)
library(fields)
library(MASS)
library(spam)
library(LatticeKrig)
```


# 1. Load Data
```{r}
# environmental and soil variables
covariate_m <- read_excel("../Data/Malawi/Malawi_grain_soil.xlsx")
colnames(covariate_m) <- c("Latitude","Longitude","Crop","Se","Zn","ph","N_pct","org","Bio1","Bio12","TIM")

# subset maize data
maize_m <- subset(covariate_m, Crop == "Maize")

# remove missing values
maize_m <- maize_m[complete.cases(maize_m[,c("ph","org","Bio1","Bio12","TIM","Zn")]),]
maize_m <- maize_m[maize_m$Zn <= 77777,]

n0_maize <- nrow(maize_m)

# kriging target locations
target_m <- read.csv("../Data/Malawi/Malawi_targets.csv")

# combine locations
location <- rbind(
  as.matrix(maize_m[,c("Longitude","Latitude")]),
  as.matrix(target_m)
)

n <- nrow(location)
```


# 2. Proximity / Adjacency Matrix
```{r}
# use LKDist to find pairwise distances within a maximum distance
dis <- LKDist(location, location, delta = 0.01, distance.type = "Euclidean")

# check average number of neighbors
(length(dis$ra) - n) / n

# uniform weights
dis$ra <- ifelse(dis$ra > 0, 1, 0)

# proximity/adjacency matrix
adj_matrix <- as.dgCMatrix.spam(spind2spam(dis))
diag(adj_matrix) <- 0
```


# 3. Laplacian Matrix
```{r}
# construct graph
graph <- graph_from_adjacency_matrix(adj_matrix, weighted = TRUE)
# check if the graph is connected
is_connected(graph)

# construct sparse Laplacian matrix
Q <- laplacian_matrix(graph, sparse = TRUE)
# check degrees
summary(as.numeric(diag(Q)))
# normalize the degree of each vertex 
Q <- Q / mean(as.numeric(diag(Q)))
# add perturbation
Q <- Q + 1e-9 * Diagonal(n)
# LDU Cholesky decomposition
K <- Cholesky(Q, perm = TRUE, super = FALSE, LDL = TRUE)
# expand the original compact representation 
K0 <- expand2(K, LDL = TRUE)
```


# 4. F Matrix
```{r}
F_maize <- Matrix(0, nrow = n0_maize, ncol = n, sparse = TRUE)
for(k in 1:n0_maize) F_maize[k,k] <- 1
```


# 5. REML Parameter Estimation via h-likelihood
```{r}
# choose zinc to analyze
y <- as.matrix(maize_m[,"Zn"])

# mean-zero design matrix x
x <- apply(maize_m[,c("Bio1","TIM")], 2, function(x) x-mean(x))

# covariates of gamma glm
z1 <- c(rep(1,n0_maize), rep(0,n))
z2 <- c(rep(0,n0_maize), diag(K0$D))

# measure convergence
ep <- 1
# number of iterations
dummy <- 1
# initial parameter value
lam0 <- 5 
lam1 <- 2
est_lam <- c(lam0, lam1)

# keep track of valid locations
valid_err_idx <- seq(1,n0_maize+n,1)
valid_H_idx <- seq(1,n0_maize+n,1)

# keep track of time used
start_time <- Sys.time()

# h-likelihood iteration
while(ep>.00001){
  # print iteration
  print(dummy)
  
  # update parameters
  lam0 <- est_lam[1]
  lam1 <- est_lam[2]
  
  # augmented model
  
  # precision matrix with parameters
  precision_matrix <- lam0*t(F_maize)%*%F_maize + lam1*Q
  
  # component of BLUP
  ZgammaZ <- rbind(cbind(lam0*t(x)%*%x,lam0*t(x)%*%F_maize), cbind(lam0*t(F_maize)%*%x,precision_matrix))
  
  # BLUP
  blup <- lam0*solve(ZgammaZ,rbind(t(x),t(F_maize))%*%y,sparse=T)
  beta_hat <- blup[1:2]
  psi_hat <- blup[-(1:2)]
  
  # error
  err <- rbind(y-x%*%beta_hat-F_maize%*%psi_hat, -K0$L1.%*%K0$P1%*%psi_hat)
  
  
  # gamma glm 
  
  # permutation matrix
  permutation <- K0$P1.*1
  # selected inverse of the precision matrix (match permutation with Cholesky decomposition)
  TD_inv_precision <- Takahashi_Davis(Q=precision_matrix, P=permutation, return_perm_chol=1)
  
  # component of hat matrix
  inv_pre_Fx <- solve(precision_matrix,lam0*t(F_maize)%*%x,sparse=T)
  A <- lam0*t(x)%*%x - lam0*t(x)%*%F_maize%*%inv_pre_Fx
  
  # H11
  H11 <- diag(lam0*(x-F_maize%*%inv_pre_Fx) %*% solve(A) %*% (t(x)-t(inv_pre_Fx)%*%t(F_maize)) + lam0*F_maize%*%TD_inv_precision$S%*%t(F_maize))
  # H22
  C <- K0$L1.%*%K0$P1%*%inv_pre_Fx%*%t(chol(solve(A)))
  H22 <- lam1*rowSums(C^2)%*%K0$D + lam1*rowSums((K0$L1.%*%K0$P1%*%TD_inv_precision$S) * (K0$L1.%*%K0$P1) * diag(K0$D))
  
  # diagonal of hat matrix
  H <- c(H11,as.array(H22))
  
  # response of gamma glm
  res <- as.matrix(err^2/(1-H))
  # valid locations for gamma glm
  valid_err_idx <- intersect(valid_err_idx, which(err>10^(-6))) 
  valid_H_idx <- intersect(valid_H_idx, which(H<1-10^(-6))) 
  valid_idx <- intersect(valid_err_idx, valid_H_idx)
  # gamma glm with default inverse link
  lam.glm <- glm( res[valid_idx]~-1+z1[valid_idx]+z2[valid_idx],family=Gamma(), weights=(1-H[valid_idx]))
  # estimation of lam0 and lam1
  lam.new <- as.numeric(lam.glm$coeff)
  
  # updating
  ep <- sum(abs(lam.new-est_lam)/est_lam)
  print(lam.new)
  est_lam <- lam.new
  dummy <- dummy+1
}

Sys.time()-start_time
```


# 6. Results and Output Files
```{r}
#------------ Results ------------# 

# BLUP
beta_hat
psi_hat
# estimated precision parameters
est_lam

# standard error of beta 
se_beta <- sqrt(diag(Takahashi_Davis(ZgammaZ))[1:2])
# standard error of lambda 
summary(lam.glm)

# standard error of prediction 
se_predict <- sqrt(diag(TD_inv_precision$S))

# log REML likelihood
2-lam.glm$aic/2


#------------Output Files------------# 

dir_path <- file.path("..", "Output")

write(psi_hat, file = file.path(dir_path, "malawi_psi_hat.txt"), ncolumns = 1)
write(se_beta, file = file.path(dir_path, "malawi_se_beta.txt"), ncolumns = 1)

write.table(location, file = file.path(dir_path, "location_m.txt"),row.names = FALSE,col.names = TRUE)
write.table(maize_m, file = file.path(dir_path, "maize_m.txt"),row.names = FALSE,col.names = TRUE)
```






